SBAT: A Simple Browser-based Annotation Tool with Interactive Visualization and Github Support

### 2024-09-13 paper/main.tex
expand introduction

%% Uncomment once you want to add the anti-plagiarism statement file
% \newpage
% \includepdf[pages=-]{name-of-anti-plagiarism-statement-file}

\newpage
\tableofcontents
%\listoftables
%\listoffigures
\newpage
\pagenumbering{arabic}

\section{Introduction}
This paper introduces SBAT, a lightweight and easy-to-use web annotation tool. It supports annotation tasks that require attaching labels to text data, such as POS tagging or labeling for named entity recognition. Various apps have been developed for the purpose of text annotation, but they often require a dedicated server or a local installation, which complicates the usage. Therefore, SBAT is designed to be an online annotation tool that is easy to access and simple to use. It is built on the Brat annotation interface, but is completely browser-based and server-free, while preserving most of the key functionalities. Additionally, it includes support for integration with Github's file management system.

@@

Among the annotation tools, Brat \cite{stenetorp-etal-2012-brat} is a powerful scheme-neutral annotation tool with an intuitive user interface. Brat can be used for tasks including but not limited to POS tagging, named entity recognition, semantic role labeling, dependency and verb-frame annotations, and enables user-defined constraints-checking. It provides a high-quality visualization of annotations by its browser-based UI component implemented in XHTML, Scalable Vector Graphics (SVG) and JavaScript. Last but not least, Brat integrats a ML-based semantic class disambiguator that outputs multiple annotations with their possibility estimates, which has been proven to increase the efficiency of annotators in several annotation projects.\\

@@

Despite the success of Brat, it is not the simplest tool to start using. A CGI-capable server is required by its installation, in order to combine all user modifications in real-time with the stored data. That is where brat-frontend-editor \cite{brat-frontend-editor}, a forked project of Brat, differentiates itself from the original project. Brat-frontend-editor is a standalone browser version of Brat that keeps some of its essential editing functions but removes its server-side code. It can be imported as a module in vanilla JavaScript, Angular or React. This project is since 2018 no longer maintained, so many dependencies have broken, and it no longer keeps up with the new updates of the original Brat. However, its transformation of server-side operations into browser-side operations serves as an important reference in the development of SBAT.\\

@@

While Brat serves as a general-purpose annotation tool, more recent tools target specific annotation schemes. UD Annotatrix \cite{tyers-etal-2017-ud}, for example, is specifically designed for annotations of universal dependencies. It accepts several input formats including plain text and CoNNL-U, and offers graphical editing functions. Unlike Brat, it offers a stand-alone module written in JavaScript with locally saved dependencies, which stores imported corpora in localStorage and allows offline usage. Otherwise, there is also a server module to save corpora on the server.\\

@@

Another specialized annotation tool is INCEpTION \cite{klie-etal-2018-inception}, which is used for semantic annotation. One of its characteristics is the extensive adoption of machine-learning models to assist the annotation process. It not only utilizes recommenders based on machine learning models to provide users with suggestions for possible labels, but also provides an active learning mode where user feedback can be used to further improve the quality of those suggestions. It also has a comprehensive user management system, where the admin can create multiple user accounts for the project and assign them different roles. In terms of the software itself, INCEpTION is a Java application with Spring Boot backend and a web user interface, so its usage requires the installation of Java as well as the software.\\

@@

Lastly, the previous version of SBAT \cite{SBAT} serves as one of the foundations for the current version. The previous version shares the same goal: to create a simple but functional annotation tool that operates entirely in the browser, without any server-side configuration. It is implemented using HTML and JavaScript, and can be used to attach user-defined labels to paragraphs. It also allows the local import and export of annotation files. Building on that, the current version addresses the shortcomings of the previous one by adding functionalities such as span annotations, annotation visualization, and Github support.

@@

@@

\section{Conclusion}
SBAT is an annotation tool for attaching labels to linguistic data, which aims to provide a simpler and no-server solution for annotation projects. It is completely browser-based and only requires a browser for usage. It retains the essential features of Brat's annotation editor, such as annotation operations and visualizations, while offering integration with Github, enabling user authentication and the import and export of annotations from and to Github repositories.\\

@@

\bibliography{bibliography}
\bibliographystyle{plain}

>>>>

%% Uncomment once you want to add the anti-plagiarism statement file
% \newpage
% \includepdf[pages=-]{name-of-anti-plagiarism-statement-file}
\newpage
\section*{\centering{Abstract}}
This paper introduces SBAT, a lightweight and easy-to-use web annotation tool. It supports annotation tasks that require attaching labels to text data, such as POS tagging or labeling for named entity recognition. Various apps have been developed for the purpose of text annotation, but they often require a dedicated server or a local installation, which complicates the usage. Therefore, SBAT is designed to be an online annotation tool that is easy to access and simple to use. It is built on the Brat annotation interface, but is completely browser-based and server-free, while preserving most of the key functionalities. Additionally, it includes support for integration with Github's file management system.
\newpage
\tableofcontents
%\listoftables
%\listoffigures
\newpage
\pagenumbering{arabic}

\section{Introduction}
In computational linguistics, annotating linguistic data has been a crucial step for many studies, based on which it is possible to examine patterns, study linguistic phenomena, train statistical models or testify linguistic theories. And among all types of linguistic data, text annotation is the most prevalent. Text annotation refers to the labeling or classifying of text data, which is then used for various purposes. For instance, the text data with named entities tags can be used to train a NER (named entity recognition) model for Natural Language Processing. But annotating is not limited to named entities, tagging semantic roles, dependencies, parts of speech, sentiments, relations and so on are all considered text annotations in common sense. Moreover, although automatic annotations by the computer achieved good performance in some cases, most annotation tasks still require the annotation or validation of human annotators, highlighting the necessity of annotation tools.\\
\\
For manual or computer-assisted creations of annotated data, a variety of applications have been developed over the years, each with a specific focus or for a specific annotation scheme. These tools can recognize spans of characters and their links, and enable the user to associate labels to those spans and links \citep{inbook}. Notable examples include BRAT for text span annotation and relation annotation, UD Annotatrix for Universal Dependencies annotation, and INCEpTION for Semantic annotation. These tools not only provide an editor-like user interface for annotating operations, but many also enable suggestions generated by machine-learning models to speed up the annotation creation and increase the accuracy. Designed for usage in annotation projects, most of them provide user management functionalities through user registration and role assignment, in order to facilitate collaboration of annotators in the annotation process. In terms of usage, many of those tools are dependent on Java or Python libraries, and therefore require installation of Java or Python prior to use. Some of them have a browser-based frontend, while others are desktop applications.\\
\\
While those existing annotation tools provide various solutions across different schemes for the conduction of annotation projects, they often require complicated configuration on either local devices or the server side \citep{yang2018yeddalightweightcollaborativetext}, which is not ideal for smaller annotation projects with limited time and low budget. Furthermore, most of them do not have an efficient version control system, which means keeping track of changes or reverting back to previous versions is either impossible or requires a lot of effort. Taking those limitations into account, SBAT is designed to be a browser-only, lightweight text span annotation tool, with integrated support for reading and writing files from and to Github repositories. By making SBAT browser-based, its only software dependency is a web browser, which is already available on every modern computer. And its support for Github repository operations opens the door to the application of Github's version management functionalities in the annotation project.\\
\\
The rest of this paper is therefore divided into four sections, each dedicated to the introduction and discussion of annotation tools and SBAT. The second chapter deals with the related work in the field, presenting a few representative annotation tools developed over the years, along with an overview of their functionalities, advantages and limitations. The third chapter introduces the main features of SBAT, such as its interactive user interface and its support for Github, backed up with detailed instructions and illustrations. Following that, the fourth chapter talks about the implementation details in the development of SBAT. Finally, this paper ends with concluding remarks about SBAT, pointing out its strength and limitations.

\newpage

@@

Among the annotation tools, BRAT \citep{stenetorp-etal-2012-brat} is a powerful scheme-neutral annotation tool with an intuitive user interface. Brat can be used for tasks including but not limited to POS tagging, named entity recognition, semantic role labeling, dependency and verb-frame annotations, and enables user-defined constraints-checking. It provides a high-quality visualization of annotations by its browser-based UI component implemented in XHTML, Scalable Vector Graphics (SVG) and JavaScript. Last but not least, Brat integrats a ML-based semantic class disambiguator that outputs multiple annotations with their possibility estimates, which has been proven to increase the efficiency of annotators in several annotation projects.\\

@@

Despite the success of BRAT, it is not the simplest tool to start using. A CGI-capable server is required by its installation, in order to combine all user modifications in real-time with the stored data. That is where brat-frontend-editor \citep{brat-frontend-editor}, a forked project of Brat, differentiates itself from the original project. Brat-frontend-editor is a standalone browser version of Brat that keeps some of its essential editing functions but removes its server-side code. It can be imported as a module in vanilla JavaScript, Angular or React. This project is since 2018 no longer maintained, so many dependencies have broken, and it no longer keeps up with the new updates of the original Brat. However, its transformation of server-side operations into browser-side operations serves as an important reference in the development of SBAT.\\

@@

While Brat serves as a general-purpose annotation tool, more recent tools target specific annotation schemes. UD Annotatrix \citep{tyers-etal-2017-ud}, for example, is specifically designed for annotations of universal dependencies. It accepts several input formats including plain text and CoNNL-U, and offers graphical editing functions. Unlike Brat, it offers a stand-alone module written in JavaScript with locally saved dependencies, which stores imported corpora in localStorage and allows offline usage. Otherwise, there is also a server module to save corpora on the server.\\

@@

Another specialized annotation tool is INCEpTION \citep{klie-etal-2018-inception}, which is used for semantic annotation. One of its characteristics is the extensive adoption of machine-learning models to assist the annotation process. It not only utilizes recommenders based on machine learning models to provide users with suggestions for possible labels, but also provides an active learning mode where user feedback can be used to further improve the quality of those suggestions. It also has a comprehensive user management system, where the admin can create multiple user accounts for the project and assign them different roles. In terms of the software itself, INCEpTION is a Java application with Spring Boot backend and a web user interface, so its usage requires the installation of Java as well as the software.\\

@@

Lastly, the previous version of SBAT \citep{SBAT} serves as one of the foundations for the current version. The previous version shares the same goal: to create a simple but functional annotation tool that operates entirely in the browser, without any server-side configuration. It is implemented using HTML and JavaScript, and can be used to attach user-defined labels to paragraphs. It also allows the local import and export of annotation files. Building on that, the current version addresses the shortcomings of the previous one by adding functionalities such as span annotations, annotation visualization, and Github support.
\newpage

@@

\newpage

@@

\newpage
\section{Conclusion}
SBAT is an annotation tool for attaching labels to linguistic data, which aims to provide a simpler and no-server solution for annotation projects. It is completely browser-based and only requires a browser for usage. It retains the essential features of Brat's annotation editor, such as annotation operations and visualizations, while offering integration with Github, enabling user authentication and the import and export of annotations from and to Github repositories.\\

@@

\newpage
\bibliography{bibliography}
\bibliographystyle{plainnat}
